A multi-scale story of the diffusion of a new technology: the web

### 2025-03-12 paper/submission2/tranos2025.tex
paper v2

@@

\subsection{System-level analysis}\label{system-level-analysis}

@@

\caption{\label{s_uk}Grwoth curve, UK}

@@

Figure \ref{s_uk} plots the cumulative adoption of the Web in total in
the UK during the 1996-2012 period. Not surprisingly, the cumulative
adoption is well represented by a curve that resembles the figure S. The
vertical line around year 2003 illustrates the point where the modelled
cumulative adoption is equal to 50\% of the maximum. This \(t_0\)
\emph{inflection} point signals the maximum adoption speed and is used
here to determine whether a UK LAD reached that inflection point earlier
or later than the UK average. Specifically, the S curve and the
inflection point are estimated for every LAD individually and then
compared against the UK average. LAD that reached their inflection point
earlier than the country average are labelled as \emph{fast} and the
rest as \emph{slow}. Figure \ref{s_map} maps this pattern. There is a
clear high concentration of fast adopting LAD in the South and East of
England as expected. However, there is also such a concentration in
Wales, Scotland and the North West of England. At a more detailed level,
there are some expected examples of LAD with relevant industrial
backgrounds delineated as fast: the City of London, a world-renowned
cluster of finance industries \citep{cook2007role}, and Reading, a town
with high-tech service industries in proximity to London and its main
airport, Heathrow \citep{england2005polynet}. However, there are also
LAD which were expected to appear as fast -- e.g.~Hackney in central
London and Bristol, a well-established creative cluster
\citep{oatley1999cultural, bassett2002cultural} -- but were delineated
as slow. Nevertheless, there is obvious spatial clustering of fast and
slow LAD. Importantly, out of the ten fastest LAD, nine were located in
the South East of England and London and one in Cambridge (see Table
\ref{table.s.lads} in the Appendix). The above observations remain
almost unchanged when the analysis includes websites with up to 10
unique postcodes -- see Figures \ref{s_uk10} and \ref{s_map10} in the
Appendix -- advocating towards the robustness of the analysis.

@@

@@

\caption{\label{s_map}LAD adoption rates}

@@

The S-curves of the cumulative adoption of Web in the UK in total and in
the UK LAD demonstrate a pattern well aligned with all previous studies
discussed in Section~\ref{sec-litreview}. Importantly, the above
findings are in line with previous studies exploring the diffusion of
web technologies \citep{PAPAGIANNIDIS2015308}, social media
\citep{lengyel2020role} and online shopping \citep{bakher2013diffusion}.
Similar to \citet{beardsell1999spatial}, who analysed the spatial
evolution of computer industry employment across 317 US metro areas,
spatial heterogeneity is evident in the diffusion speed when the scale
of analysis is as detailed as the LAD. Importantly, no previous studies
estimated such cumulative adoption curves for areal units as small as
the UK LADS.

The next step assesses the stability and volatility of the LAD in terms
of their adoption of web technologies. As highlighted in the literature
\citep{risk_perceptions}, different agents have different perceptions
about and levels of acceptance of the risks and the potential economic
returns associated with the adoption of new technologies -- see for
instance the seminal work of \citet{venkatesh2000theoretical}. To reveal
such aggregated patterns, Figure \ref{rank} plots the ranking of UK LAD
based on website density. To decrease noise, the average ranking of
1996-1998 and 2010-2012 is plotted instead of the individual years. The
colours depict the 5th and 95th percentile of the absolute difference of
the LAD ranking between 1996-1998 and 2010-2012. While there are quite a
few obvious cases of LAD that maintained their position between the
beginning and the end of the study period both at the top or at the
bottom of the hierarchy, there are also quite a few LAD that changed
drastically their position. Some of these LAD enjoyed a process that at
the first instance looks like \emph{leapfrogging} since they managed to
jump at the top of the hierarchy despite their slow start. There is
extensive literature regarding the potential benefits of technological
leapfrogging. The underpinning argument is that latecomers can adopt and
benefit from new technologies that have been developed elsewhere without
incurring the hefty initial R\&D costs \citep{teece2008firm}. Although
the leapfrogging literature does not pay much attention to cities and
regions \citep{yu2018sustainability}, previous research highlighted the
long term and sustained productivity benefit of the early adopters of
web technologies in the UK \citep{tranosuk}. So, although the LAD system
is volatile, it is not clear whether the LAD with high concentration of
late adopters will gain any latecomer-related benefits in a way similar
to countries experiencing technological leapfrogging. Still, the
opportunity to catch up that \citet{perkins2005international}
illustrated at the country level, is also visible at the much more
detailed LAD scale. As previously, this is consistent with including
websites with up to 10 postcodes (see Figure \ref{rank10}).

\caption{\label{rank}Dynamics of wed diffusion (LAD)}

\subsection{Exploratory analysis}\label{exploratory-analysis}

Figures \ref{morani} and \ref{lisa} offer a first insight into whether a
neighbourhood mechanism underpins the diffusion of the Web in the UK as
they present the Moran's I and the LISA maps of website density
respectively for LAD and OA. Starting from the former, spatial
autocorrelation was higher in the beginning of the study period and then
after 2000 dropped slightly and stabilised around 0.2. This reflects the
early concentration of high website density around London, which over
time diffused as high website density clusters can be seen in other
parts of the country away from London (Figure \ref{lisa}). An almost
reverse pattern can be observed for OA. At the beginning of the study
period Moran's I was around 0.5 and it plateaued after 2000 around 0.2.
Because of the very small size of OA, at the early stages of the
diffusion of web technologies their adoption was spatially scattered.
This is reflected in the lack of any significant clusters in 1996
(Figure \ref{lisa}). Eventually, as the adoption rate increased, more
such clusters of high website density were formed and this is reflected
both in the Moran's I and the LISA maps in Figures \ref{morani} and
\ref{lisa}. A similar pattern is observed for the extended dataset (see
Figures\ref{morani10} and \ref{lisa10}). In this case, the magnitude of
spatial autocorrelation is slightly higher illustrating a stronger
spatial clustering mechanism. All in all, the exploratory spatial data
analysis advocates towards an underpinning neighbourhood mechanism and
the different scales of analysis illustrate how it evolved differently
over time. Interestingly, the magnitude of Moran's I is similar to the
ones revealed by \citet{ding2010modeling} about mobile phone adoption in
China albeit the much more detailed spatial scale of this paper.

@@

{\centering \includegraphics[width=1\textwidth,height=0.8\textheight]{tranos2025_files/figure-pdf/unnamed-chunk-6-1.pdf}

@@

\subsection{Modelling outputs}\label{modelling-outputs}

The next section incorporates the above mechanisms of the spatial
diffusion of the Web into a modelling framework. The aim is to use
variables depicting these spatial processes in order to predict the
diffusion of the Web as a new technology in the UK over space and time
and across different scales. Specifically, four different models are
estimated. Firstly, all the data points for the OA and LAD are utilised
in order to build two RF models and assess their capacity to predict the
adoption of the Web at the two different scales. These two models will
reveal the predictive capacity of the diffusion mechanisms and also
demonstrate how the importance of such variables changes across
different scales. The next two sets of models are again trained at these
two scales: OA and LAD. However, instead of using all the data points,
the OA and the LAD from one of the twelve UK regions are held out for
the model training. Then, the trained model is used to predict website
density for the OA or the LAD of the held-out region. This process takes
place recursively for all twelve UK regions. The regional differences in
the predictive capacity of the different samples will reveal how
dissimilar these spatial process are across regions and, importantly, at
different scales.

It needs to be highlighted here that the cross validation (CV) for all
models is spatially and temporally sensitive. Instead of using 10 random
and space- and time-agnostic folds,the \texttt{CAST} package is employed
which allows holding back data points from specific years and spatial
units and use them for testing in order to estimate the model
performance \citep{meyer2018improving}.

The models include variables that capture the three processes that the
relevant literature and the descriptive analysis highlighted. Namely,
the models capture: (i) a hierarchical mechanism with diffusion running
from main centres to secondary ones, (ii) a neighborhood mechanism
according to which diffusion first hits nearby locations, and (iii) the
rather canonical pattern of diffusion over time as reflected in the
S-shaped pattern in the cumulative level of adoption.

To capture the hierarchical mechanism the models include as predictors a
one year lag of website density in London, the largest city in the UK, a
one year lag of the website density in the nearest city and the same for
the nearest retail centre. Due to the small sizes of the retail centres,
the latter is only relevant for and included in the OA-level models. In
addition, the models include the distance to London, the nearest city
and the nearest retail centre. The underlying logic is that the level of
website adoption in a spatial unit depends on the level of the adoption
in places further up in the urban hierarchy the previous year. The
inclusion of the distance variables incorporates spatial structure into
the hierarchy argument. To depict the neighbourhood mechanism, the
website density of the neighbouring spatial units in the previous year
is employed. This is calculated using rook continuity. Again, the
underpinning rationale is that the level of web adoption within a
spatial unit depends on the level of web adoption in the neighbouring
spatial units the year before. This represents the `hitting nearby
locations first' argument. Therefore, the spatial and temporal lag of
the website density in LAD and OA is calculated. Lastly, the temporal
pattern of the cumulative adoption is captured by a time trend variable.
Hence, all four models will follow the following generic form (Eq.
\ref{model}):

\begin{align} \label{model}
Website\,Density_{t} \sim Distance\,London +
Website\,density\,London_{t-1} +\notag\\
Distance\,Nearest\,City +
Website\,density\,Nearest\,City_{t-1} +\notag\\
Distance\,Nearest\,Retail_{i} +
Website\,density\,Nearest\,Retail_{t-1} +\notag\\
W*\, Website\,density_{t-1} +\notag\\ 
year_{t}
\end{align}

Table~\ref{tbl-model-metrics} presents the model performance for the
first set of models, for which all data points are employed for training
and testing via CV. The first one is trained and tested on 374 LAD and
the second on 232,296 OA, both for the 16 year period (1997-2012). The
results are remarkably good considering that the are the outcome of
space and time sensitive CV, so the the model does not suffer from
overfitting. At the LAD level the model predicts 85\% of the variation
of website density. Both error metrics indicate that the model error is
\((\frac{1}{20}, \frac{1}{30})\) of a website per firm. At the OA the
\(R^2\) drops down to 32\%. Considering its granularity, this is still a
remarkable performance. To contextualise it, the model results in a
\(MAE\) equal to one website for areas small enough to host less than
140 households. \footnote{According to the Office for National
  Statistics, 80\% of OA in England and Wales host 110-139 households,
  \href{https://www.ons.gov.uk/census/2001censusandearlier/dataandproducts/outputgeography/outputareas}{www.ons.gov.uk}.}
Because of the small size of the spatial units, the distribution is
highly skewed and a significant part of them is not linked to any
websites. In 1997 only 1\% of the UK OA were associated with at least
one website. This should not come as a surprise as this was the very
beginning of the commercial Internet and any activities with a digital
footprint were concentrated in a handful of areas. This was clearly
illustrated in Figure \ref{lisa}. At the end of the study period almost
half of the UK OA were not associated with a website. Again, given the
granularity of the data this should not come as a surprise.

\begin{longtable}[]{@{}llll@{}}
\caption{Model metrics}\label{tbl-model-metrics}\tabularnewline
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Local Authorities & 0.028 & 0.850 & 0.019 \\
Output Areas & 3.284 & 0.320 & 1.034 \\
\end{longtable}

The results in Table~\ref{tbl-model-metrics} are robust against
different specification. As noted earlier, the data imputation process
resulted in a modest improvement of the predictive capacity of the
model. As per Table~\ref{tbl-model-metrics-no-correction} all accuracy
metrics are only slightly worst when the data imputation is not applied.
As expected, the improvement is higher for the more detailed scale of
analysis. Then, Table~\ref{tbl-model-metrics-10} presents the model
results for the extended dataset, which includes websites with up to 10
unique postcodes. While the results for the LAD are slightly less
accurate, the \(R^{2}\) of the OA model is more than double in
comparison to the base results in Table~\ref{tbl-model-metrics}. All in
all, the modelling exercise revealed how well can the three diffusion
mechanisms predict the diffusion of the Web as a new technology.
Importantly, this predictive power is highly robust against against
different scales and delineations of the diffusion of the web.

\begin{longtable}[]{@{}llll@{}}
\caption{Model metrics without data imputation
}\label{tbl-model-metrics-no-correction}\tabularnewline
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Local Authorities & 0.032 & 0.810 & 0.019 \\
Output Areas & 5.000 & 0.205 & 1.047 \\
\end{longtable}

\begin{longtable}[]{@{}llll@{}}
\caption{Model metrics for the extended
dataset}\label{tbl-model-metrics-10}\tabularnewline
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Local Authorities & 0.188 & 0.77 & 0.122 \\
Output Areas & 22.467 & 0.478 & 5.681 \\
\end{longtable}

Figure \ref{var.imp} plots the importance of the different predictors.
When the focus is on LAD, the website density in the nearest city, in
London and in the neighbouring LAD the year before are the most
important predictors. They are followed by the yearly trend, while the
spatial configuration as reflected in distances to London or the nearest
city only plays a minor role. This can be attributed to the relatively
coarse spatial scale of LAD. Nevertheless, all previously discussed
spatial processes are at play in the diffusion of web technologies at
the LAD level: the first two predictors depict the hierarchical
mechanism, the spatial and temporal lag of website density the
neighbourhood mechanism, and the yearly trend the time-sensitive
cumulative adoption pattern.

When the much more granular scale of OA is adopted, the picture is
almost reversed. The two most important predictors are the distance to
London and to the nearest city followed by the spatial and temporal lag
of website density and the distance to the nearest retail centre. They
still depict a hierarchical mechanism, but proximity to the different
population centres is more important than their lagged web densities in
predicting website diffusion. The neighbourhood mechanism is still
strong but less important at this scale. What is interesting is the
almost negligible role of the yearly trend and London's website density.
While the former probably illustrates the large heterogeneity in how the
Web has been adopted at this very fine scale, the latter highlights that
the importance of past web adoption rates in large population centres is
surpassed by proximity to them and spatial configuration at this scale.

The above pattern of variable importance is largely in accordance with
the models estimated for the extended dataset. A comparison with Figure
\ref{var.imp10} illustrates that the main difference is the importance
of the neighbourhood mechanism for the OA as the space and time lag of
the website density is by far the most important predictor when the
extended dataset is employed. But still, the variable importance
patterns for the two datasets are not dissimilar.

@@

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Regional differences\label{table.regions}}\tabularnewline
\toprule\noalign{}
Region & \(R^2\) LAD & Rank LAD & \(R^2\) OA & Rank OA \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Region & \(R^2\) LAD & Rank LAD & \(R^2\) OA & Rank OA \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
South East & 0.945 & 1 & 0.116 & 8 \\
West Midlands & 0.922 & 2 & 0.125 & 4 \\
Wales & 0.915 & 3 & 0.125 & 5 \\
Yorkshire and The Humber & 0.909 & 4 & 0.160 & 2 \\
East Midlands & 0.903 & 5 & 0.120 & 7 \\
North East & 0.896 & 6 & 0.124 & 6 \\
South West & 0.894 & 7 & 0.184 & 1 \\
London & 0.881 & 8 & 0.093 & 9 \\
East of England & 0.874 & 9 & 0.092 & 10 \\
North West & 0.837 & 10 & 0.143 & 3 \\
Scotland & 0.780 & 11 & 0.055 & 12 \\
Nortern Ireland & 0.579 & 12 & 0.060 & 11 \\
\end{longtable}

Table \ref{table.regions} presents the results of the recursive hold out
models and allows to (i) highlight the regional heterogeneity of the
spatial processes behind the diffusion of the Web, and (ii) compare this
heterogeneity across the two different scales of analysis. It
illustrates some striking similarities, but also a few significant
differences. At the top of the similarity with the rest of the country
we see West Midlands and Yorkshire and the Humber for both scales. The
web density for the LAD and OA of these regions can be very well
predicted by models trained on data for the LAD and OA of the rest of
the country. In other words, the spatial diffusion mechanisms for these
regions represent quite well the UK's average. This should not come as
as surprise given their urban structure: they both contain a large urban
centre (Birmingham and Leeds), smaller commuter towns as well as more
rural settlements. The bottom of the similarity ranking is populated by
Scotland, Northern Ireland, East of England and London for both scales.
The diffusion mechanisms for these regions diverge significantly from
the country's average and this could be attributed to rurality and
remoteness or to high levels of urbanisation and sophistication of
economic activities for the case of London and for East of England,
which contains the Cambridge area.

Still, there are three regions the spatial diffusion mechanisms of which
differ a lot between the two scales. North West and South West are very
close to the UK's average at the more granular OA scale, while the
opposite applies for South East. Interestingly, the first two regions
are well-known tourist destinations (Lake District and Cornwall-Devon)
and the exploratory analysis revealed clusters of high web density only
visible at the OA scale (Figure \ref{lisa}. Then, the South East is very
well integrated with the Greater London, hosts the broader Oxford area
-- a known hub of research excellence -- and Figure \ref{lisa} revealed
high web density clusters at the LAD level.

>>>>

It needs to be highlighted here that the cross validation (CV) for all
models is spatially and temporally sensitive. Instead of using 10 random
and space- and time-agnostic folds, the \texttt{CAST} package is
employed which allows holding back data points from specific years and
spatial units (regions) and use them for testing in order to estimate
the model performance \citep{meyer2018improving}.

An alternative specification could model the growth of website density.
However, such a strategy is problematic because of the large frequency
of OA without websites, which is to be expected given their small size
and high granularity. Still, as a robustness text, the results of a
growth model for LAD are presented in the Appendix.

Lastly, to test the second hypothesis (H\textsubscript{2}) about the
early adoption and the latecomer effects, a different research framework
is developed. Firstly, the S-curve and the estimated \(t_0\) for every
LAD are employed and mapped in order to depict the spatial heterogeneity
and specific patterns of the speed of the diffusion of the Web. Then,
the volatility and the stability of the diffusion across the different
LAD is analysed using tools from the complex systems domain
\citep{Batty2009} namely changes in rankings of LAD based on website
density. Both the S-curve estimation and the volatility analysis focus
only on LAD as the very large number of OA would have made such analysis
difficult to visualise and interpret.

@@

\subsection{The three diffusion mechanisms at granular geographical
scales}\label{the-three-diffusion-mechanisms-at-granular-geographical-scales}

Figure \ref{s_uk} plots the cumulative adoption of the Web in the UK as
a whole during the 1996-2012 period. In line with the theoretical
expectations discussed in the previous section, the cumulative adoption
of the Web is well represented by a curve that resembles the letter S.
The vertical line in the beginning of 2003 (2003.158) illustrates the
point when the modelled cumulative adoption was equal to 50\% of the
maximum. This \(t_0\) point is known as the \emph{inflection} point and
signals the maximum adoption speed. The results remain unchanged when
the extended subset of websites, which include up to 10 unique postcodes
is used (see Figure \ref{s_uk10} in the Appendix). Importantly though,
the extended subset of websites indicates a saturation level close to
100\% at the end of the study period.

@@

\caption{\label{s_uk}Cumulative adoption of the Web, UK}

@@

What escaped from previous studies was εexploring whether such a pattern
also describes cumulative adoption of new technologies at sub-national
scales. To test this, the same logistic function is estimated for each
of the 374 LAD. The detailed results are presented in Table
\ref{table.s.lads} in the Appendix. The key finding is that for 342 LAD
(91\% of all LAD) the \(R^2 > 0.9\) for the logistic function, which
indicates that even at local scale the cumulative adoption of the Web is
characterised by a pattern similar to the one observed for countries.
Graphically, all these LAD curves are plotted in Figure \ref{s_lad_uk}.
The different colour lines illustrate whether a LAD reached \(t_0\)
faster or slower than the UK in total. This heterogeneity is further
discussed later in this section.

This is the first time that such cumulative adoption curves have been
estimated for areal units as small as the UK LAD. Understanding and
modelling the dynamics of technological diffusion and, ultimately, being
able to predict diffusion rates at such small local scales can directly
support the deployment of infrastructural networks (e.g.~5G) and local
technology policies. The above are in line with previous studies
exploring the diffusion of web technologies
\citep{PAPAGIANNIDIS2015308}, social media \citep{lengyel2020role} and
online shopping \citep{bakher2013diffusion}.

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{../../outputs/s/lad_per_firm_all.png}

@@

\caption{\label{s_lad_uk}Cumulative adoption of the Web, LAD}

@@

To assess the second diffusion mechanisms, Figures \ref{morani} and
\ref{lisa} offer a first insight into whether a neighbourhood or, in
other words, a distance mechanism underpins the diffusion of the Web in
the UK as they present the Moran's I and the LISA maps of website
density respectively for LAD and OA. Starting from the former, spatial
autocorrelation was higher in the beginning of the study period and then
after 2000 dropped slightly and stabilised around 0.2. This reflects the
early concentration of high website density around London, which over
time diffused as high website density clusters can be seen in other
parts of the country away from London (Figure \ref{lisa}). An almost
reverse pattern can be observed for OA. At the beginning of the study
period Moran's I was around 0.5 and it plateaued after 2000 around 0.2.
Because of the very small size of OA, at the early stages of the
diffusion of web technologies their adoption was spatially scattered.
This is reflected in the lack of any significant clusters in 1996
(Figure \ref{lisa}). Eventually, as the adoption rate increased, more
such clusters of high website density were formed and this is reflected
both in the Moran's I and the LISA maps in Figures \ref{morani} and
\ref{lisa}. A similar pattern is observed for the extended dataset (see
Figures\ref{morani10} and \ref{lisa10}). In this case, the magnitude of
spatial autocorrelation is slightly higher illustrating a stronger
spatial clustering mechanism. All in all, the exploratory spatial data
analysis advocates towards an underpinning neighbourhood mechanism,
which is present at different scales. Interestingly, the magnitude of
Moran's I is similar to the ones revealed by \citet{ding2010modeling}
about mobile phone adoption in China albeit the much more detailed
spatial scale of this paper.

@@

{\centering \includegraphics[width=1\textwidth,height=0.8\textheight]{tranos2025_files/figure-pdf/unnamed-chunk-5-1.pdf}

@@

After discussing the descriptive findings, the validity of the above
diffusion mechanisms at the local scale is assessed with a predictive
modelling framework described in Section~\ref{sec-datamethods}. The
reader is reminded that four models are estimated: for the OA and LAD,
firstly all data points are employed to assess the capacity of the three
mechanisms to to predict the adoption of the Web at these two scales;
then, the LAD/OA from one of the twelve regions are recursively held out
to assess the spatial heterogeneity of these mechanisms across the
different UK regions.

Table~\ref{tbl-model-metrics} presents the model performance for the
first set of models, for which all data points are employed for training
and testing via CV. The first one is trained and tested on 374 LAD and
the second on 232,296 OA, both for the 16 year period (1997-2012). The
results are remarkably good considering that the are the outcome of
space and time sensitive CV, so the the model does not suffer from
overfitting. At the LAD level the model predicts 83\% of the variation
of website density. Both error metrics indicate that the model error is
fraction of a website per firm (0.020 and 0.029). At the OA the \(R^2\)
drops down to 33\%. Considering its granularity, this is still a
remarkable performance. To contextualise it, the model results in a
\(MAE\) equal to 1.2 websites for areas small enough to host 40-250
households. Because of the small size of the spatial units, the
distribution is highly skewed and a significant part of them is not
linked to any websites. In 1997 only 1\% of the UK OA were associated
with at least one website. This should not come as a surprise given that
this was the very beginning of the commercial Internet and any
activities with a digital footprint were concentrated in a handful of
areas, as illustrated in Figure \ref{lisa}. At the end of the study
period almost half of the UK OA were not associated with a website.
Again, given the granularity of the data this was expected.

\begin{longtable}[]{@{}llll@{}}
\caption{Model metrics}\label{tbl-model-metrics}\tabularnewline
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Local Authorities & 0.029 & 0.824 & 0.020 \\
Output Areas & 3.218 & 0.333 & 1.233 \\
\end{longtable}

The results in Table~\ref{tbl-model-metrics} are robust against
different specifications. As noted earlier, the data imputation process
resulted in a modest improvement of the predictive capacity of the
model. As per Table~\ref{tbl-model-metrics-no-correction}, all accuracy
metrics are only slightly worst when the data imputation is not applied.
As expected, the improvement is higher for the more detailed scale of
analysis. Then, Table~\ref{tbl-model-metrics-10} presents the model
results for the extended dataset, which includes websites with up to 10
unique postcodes. While the results for LAD are only slightly less
accurate, the \(R^{2}\) of the OA model is much higher in comparison to
the base results in Table~\ref{tbl-model-metrics}. In addition,
Table~\ref{tbl-model-growth-metrics} in the Appendix presents the
accuracy results of a LAD model, the dependent variable of which is the
yearly growth rate of website density. Even for such a specification
(see Equation \ref{model.growth} in the Appendix) the predictive
capacity of the model is exceptionally high (\(R^2 = 0.5\)). All in all,
the modelling exercise revealed how well the three diffusion mechanisms
predict the diffusion of the Web as a new technology. Importantly, this
predictive power is highly robust against against different scales and
delineations of the diffusion of the web.

\begin{longtable}[]{@{}llll@{}}
\caption{Model metrics without data
imputation}\label{tbl-model-metrics-no-correction}\tabularnewline
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Local Authorities & 0.032 & 0.781 & 0.021 \\
Output Areas & 4.644 & 0.244 & 1.237 \\
\end{longtable}

\begin{longtable}[]{@{}llll@{}}
\caption{Model metrics for the extended
dataset}\label{tbl-model-metrics-10}\tabularnewline
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
& \(RMSE\) & \(R^{2}\) & \(MAE\) \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Local Authorities & 0.190 & 0.749 & 0.129 \\
Output Areas & 21.591 & 0.455 & 6.845 \\
\end{longtable}

Figure \ref{var.imp} plots the importance of the different predictors.
When the focus is on LAD, the website density in the neighbouring LAD,
in the nearest city and in London the year before are the most important
predictors. They are followed by the yearly trend, while the spatial
configuration as reflected in distances to London or to the nearest city
only plays a minor role. This can be attributed to the relatively coarse
spatial scale of LAD. Nevertheless, all previously discussed spatial
processes are at play in the diffusion of the Web at the LAD level: the
first two predictors depict the hierarchical mechanism, the spatial and
temporal lag of website density the neighbourhood mechanism, and the
yearly trend the time-sensitive cumulative adoption pattern.

At the much more granular scale of OA, the picture is almost reversed.
The two most important predictors are the distance to London and to the
nearest city followed by the spatial and temporal lag of website density
and the distance to the nearest retail centre. They still depict a
hierarchical mechanism, but proximity to the different population
centres is more important than their lagged web densities in predicting
website diffusion. The neighbourhood mechanism is still strong but less
important at this scale. What is interesting is the almost negligible
role of the yearly trend and London's website density. While the former
probably illustrates the large heterogeneity in how the Web has been
adopted at this very fine scale, the latter highlights that the
importance of past web adoption rates in large population centres is
surpassed by proximity to them and spatial configuration at this scale.

The above pattern of variable importance is largely in accordance with
the models estimated for the extended dataset. A comparison with Figure
\ref{var.imp10} illustrates that the main difference is the importance
of the neighbourhood mechanism for the OA as the space and time lag of
the website density is by far the most important predictor when the
extended dataset is employed. For the LAD, the importance of the
distance to the nearest city increases in favour to its website density
the previous year. But still, the variable importance patterns for the
two datasets are not dissimilar.

@@

Table \ref{table.regions} presents the results of the recursive hold out
models and allows to (i) highlight the regional heterogeneity of the
spatial processes behind the diffusion of the Web, and (ii) compare this
heterogeneity across the two different scales of analysis. It
illustrates both regional similarities and differences. At the top of
Table \ref{table.regions} the regions of South East and the West
Midlands have the highest similarity with the rest of the country. In
other words, the web density for the LAD of these regions can be very
well predicted by models trained on data from the LAD from the rest of
the country and, therefore, the spatial diffusion mechanisms for these
regions represent quite well the UK's average. This should not come as a
surprise as the spatial structure of both regions resembles the broader
UK: both regions contain some highly urbanised LAD (South East is well
integrated with the Greater London and West Midlands is the home of the
city of Birmingham), but also smaller commuter towns as well as more
rural settlements. At the other end of the spectrum, the spatial
diffusion mechanisms for Northern Ireland, Scotland and North West
diverge the most from the country average. Rurality and remoteness seem
to be common attributes for these regions. This hierarchy though is not
consistent across scales. For the more granular OA models, the spatial
diffusion mechanisms for South West, Yorkshire and The Humber and North
West regions are closer to the country average. Interestingly, these
regions are well-known tourist destinations (Lake District, Peak
District and Cornwall-Devon) and the exploratory analysis revealed
clusters of high web density only visible at the OA scale (Figure
\ref{lisa}). Then, Scotland, Northern Ireland and London have the lowest
similarity with the country average. Again, rurality, peripherality, but
also the highest level or urbanisation make the spatial diffusion
mechanisms of these regions more distinct.

\begin{longtable}[]{@{}lrrrr@{}}
\caption{Regional differences\label{table.regions}}\tabularnewline
\toprule\noalign{}
Region & \(R^2\) LAD & Rank LAD & \(R^2\) OA & Rank OA \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Region & \(R^2\) LAD & Rank LAD & \(R^2\) OA & Rank OA \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
South East & 0.952 & 1 & 0.122 & 6 \\
West Midlands & 0.926 & 2 & 0.126 & 4 \\
Wales & 0.909 & 3 & 0.122 & 8 \\
Yorkshire and The Humber & 0.908 & 4 & 0.160 & 2 \\
East Midlands & 0.904 & 5 & 0.122 & 7 \\
South West & 0.894 & 6 & 0.185 & 1 \\
North East & 0.887 & 7 & 0.124 & 5 \\
London & 0.880 & 8 & 0.090 & 11 \\
East of England & 0.876 & 9 & 0.106 & 9 \\
North West & 0.837 & 10 & 0.142 & 3 \\
Scotland & 0.774 & 11 & 0.059 & 12 \\
Nortern Ireland & 0.590 & 12 & 0.101 & 10 \\
\end{longtable}

Comparing the above with Table \ref{table.regions.10} in the Appendix,
which presents the results for the the extended dataset with websites
with up to 10 unique postcodes, the results do not differentiate much,
signalling the robustness of the analysis. The most salient difference
has to do with Northern Ireland. According to the extended sample at the
granular OA level, Northern Ireland's spatial diffusion mechanisms are
closer to the UK's average than the results based on the websites with a
unique postcode. While the latter are websites and, consequently,
economic activities solely based in Northern Ireland, websites with
multiple postcodes also include websites with postcodes in Britain and,
therefore, can be characterised by slightly different spatial diffusion
processes.

\subsection{Early adoption and
latecomers}\label{early-adoption-and-latecomers}

To test the second hypothesis (H\textsubscript{2}), the spatial
heterogeneity of Web adoption speeds is firstly analysed. As highlighted
in the literature \citep{risk_perceptions}, different agents have
different perceptions about and levels of acceptance of the risks and
the potential economic returns associated with the adoption of new
technologies -- see for instance the seminal work of
\citet{venkatesh2000theoretical}. To reveal such patterns, Figure
\ref{s_map} maps whether LAD reached their inflection point earlier or
later than the UK in total and based on this are labelled as \emph{fast}
or \emph{slow} accordingly. Inflection points have been used in the
relevant literature to compare diffusion processes across different
settings and technologies \citep{cherp2021national, woo2017exploring} as
they can enable more insightful comparisons than, for example, the
average growth over the 1996-2012 study period.

The map illustrates a clear high concentration of fast adopting LAD in
the South and East of England, as expected. There are also some
concentrations in Wales, Scotland and the North West of England. At a
more detailed level, there are some expected cases of LAD with relevant
industrial backgrounds delineated as fast: the City of London, a
world-renowned cluster of finance industries \citep{cook2007role}, and
Reading, a town with high-tech service industries in proximity to London
and its main airport, Heathrow \citep{england2005polynet}. However,
there are also LAD which were expected to appear as fast -- e.g.~Hackney
in central London and Bristol, a well-established creative cluster
\citep{oatley1999cultural, bassett2002cultural} -- but are delineated as
slow. Similar to \citet{beardsell1999spatial}, who analysed the spatial
evolution of computer industry employment across 317 US metro areas,
spatial heterogeneity is evident in the diffusion speed when the scale
of analysis is as detailed as the LAD.

Importantly, there is obvious spatial clustering of fast and slow LAD.
Out of the ten fastest LAD based on when they reached their inflection
point (see detailed Table \ref{table.s.lads} in the Appendix), nine were
located in the South East of England and London and one in Cambridge.
The above observations remain almost unchanged when the analysis
includes websites with up to 10 unique postcodes -- see Figure
\ref{s_map10} in the Appendix -- demonstrating the robustness of the
analysis. Given (i) the expectation that early adopters of new
technologies are often rewarded
\citep{rogers2010diffusion, griliches1957, ding2010modeling}, and (ii)
the previous findings illustrating the long term and sustained
productivity benefit for the early adopters of Web in the UK
\citep{tranosuk}, the observed spatial clustering of LAD with fast
diffusion rates can further reinforce existing and rooted patterns of
unequal economic productivity in the UK \citep{mccann2022productivity}.

\caption{\label{s_map}LAD Web diffusion speed, 1996-2012}
